Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2017-06-21 2017-Xudong-SMT/submit-ICTAC2017.tex
16 pages versin - submitted

In this paper, we use an example inspired from LOTOS, to illustrate
the main features of pNets and the way we deal with them in the symbolic
semantics, and in the satisfiability checking. 
%% \QIN{
%% Meanwhile, 
%% the formula using operators from CCS can also be encoded into a pNet and 
%% conducted with the same things.
%% }
%% \subsection{LOTOS Formula Example}
We choose here the expression \texttt{P>>(acc(v);Q)}, that includes the
Enable and Prefix 
operators of the LOTOS language. For the readers not familiar with
Lotos, the pNet encodings here should be sufficient to understand the
Enable operator semantics.
In Fig.~\ref{schema:lotos-pnet}, we show the encoding of the formula using pNets. 
This pNet is composed of: a toplevel node PN1 encoding the Enable
operator, a pLTS C1 acting as the {\sl controller} of the node; a hole
representing the 
process variable P; another node PN2 representing the Prefix
operator, with its controller C2, and its hole Q.

@@

% \TODO{Relate to the algorithm, or even move to section 5}
When building an Open Automaton as the semantics of a pNet, its
\emph{states}, and the shape of the predicates in its
\emph{transitions} have a specific structure:

\paragraph{States of open pNets:}\label{def-states}
  A state of an open pNet is a tuple (not necessarily finite) of the
  states of its leaves (in which we denote tuples
  in structured states as $\triangleleft\ldots\triangleright$ for better readability).

  For any pNet p, let $\overline{Leaves} = \mylangle S_i,{s_i}_0, \to_i\myrangle^{i \in L}$ be the set of pLTS at its leaves,
  then $States(p) = \{\triangleleft s_i^{i\in L}
  \triangleright| \forall i\in L. s_i \in S_i\}$.
A pLTS being its own single leave:
  $States(\mylangle S,s_0, \to\myrangle) = \{\triangleleft s \triangleright| s \in S\}$.

The initial state is defined as:
$InitState(p) = \triangleleft {{s_i}_0}^{i\in L}  \triangleright$.

@@

\paragraph{Predicates:}
Let
$\mylangle\overline{\pNet},\overline{S},\symb{SV}_k^{k\in K} \myrangle$
be a pNet. Consider a synchronization vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\[\MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Leftrightarrow
\begin{array}{l}
\exists {(a'_i)}^{i\in I},
{(b'_j)}^{j\in J},v'.\, SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
\\~~\land
\forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{array}\]
In any other case (if the action families do not match or if there is no valuation of
variables such that the above formula can be ensured) the predicate is undefined.

@@

\begin{figure}[h]
  \vspace{-10mm}

@@

\section{Implementation}
\label{section:implementation}
The VerCors platform uses (closed) pNets as the intermediate language
for some high-level language or graphical formalism to be translated
into both input for a 
model-checker and for generating executable code automatically \cite{HKM-FASE16}.
% to ease the
% specification from non-expert users, compiling it
%from those language or graphical formalism in the pNet API.

@@

\State Initialize sets $U$, $E$ for unexplored/explores global states
$U$, $L$ for result OTs;
\While{!isEmpty(U)}
	\State Extract a global state $S$ from the $U$;
	\State Recursively applying Tr1 or Tr2 on the root node and its subnets:
	\If{Subnet is pLTS}
                \State $s$ is the local state of the pLTS in the global state $S$;
		\For{each transition $tr$ starting from $s$}
			\State apply Tr1 to build the corresponding
                        $OT$; store $OT$ in $L$;
		\EndFor
	\EndIf
	\If{Subnet is pNet, apply the Tr2} 
		\For{each $sv \in SV$}
			\State Invoke $Combining()$ to generate the combination of sub-transitions;
			\State Invoke $Matching()$ to match the combination with the $sv$;
			\State Store the result set of open transitions into $L$;
		\EndFor
	\EndIf
	\For{each target global state $T$ from $OT \in L$}
		\IfThen{$(!U.contains(T))\ \&\&\ (!E.contains(T))$}{Add $T$ into the $U$;}
	\EndFor
\EndWhile
\State Prune the open transitions in $L$ using SMT solver;
\State \Return $L$;

@@

\QIN{
For the open pNets, in previous work, we already have a sketch for computing the open automaton.
The computation repeats applying Tr1 and Tr2 to deduct all the possible open transitions and pruning the 
unsatisfiable transitions by SMT solver until all the global states are explored. 
In the end, a reachability check is supposed to potentially prune the state space. 
Algorithm 1 is what we used in the implementation, there are much more details added based on 
the sketch, such as how we apply the Tr1 or Tr2.
}

In our prototype we start directly with open pNets expressed
programmatically, using the pNet API (graphical editors may be added
later). 
%From these, we compute
%the open automaton of an open pNet. We explain here our main
%implementation choices.
%:
%\begin{itemize}
%  \item Steps 2a-2b are merged, applying Tr2 from the
%premisses OTs from subnets to generate OTs without
%explicitely constructing the deduction tree.
%  \item We define a general naming schema for implementing the fresh 
%and clone functions with more structure than a simple global name
%generator. It ensures unicity, but builds names
%readable enough for debugging purpose.
%  \item Besides that, we also have designed a management of state variables and
%    assignments throughout the whole computation.
%%%   \item Satisfiability check is done only at
%%% top level of the open-transition construction. An alternative would be
%%% to submit the satisfiability check to the SMT solver at each level of
%%% the hierarchical OT construction, potentially reducing the overall number of
%%% combinations. But the submission to the SMT engine is costly, and more
%%% complexity analysis is required before deciding if this would be
%%% worthwhile. 
%\end{itemize}
%
%In the following sections, we detail:
%\begin{itemize}
%\item The structure of fresh variable names.
%\item The details of the implementation for step 2a-2b, refined into:
%  \subitem combining all possibilities of subnets activity/inactivity,
%  \subitem matching each combination with synchronization vectors.
%\item Management of local variables of pLTSs.
%\item Pruning unsatisfiable OTs using the Z3 SMT solver.
%\end{itemize}

@@

\subsection{Combining and Matching using TR2}
% \TODO{Xudong: move here a (new version) of the alogorithm loplevel sketch}
% \QIN{[Move to the head of Section 5.]}

\QIN{
The detail about the application of the Tr1 and Tr2 have not considered in the sketch, but it is an  
important part in the implementation of open transition generation. 
Each time we apply Tr1, there is not too much predicate terms, so do the other parameters, needed 
to be generated. 
When it comes to applying Tr2, the computation becomes complicated because it doing some composition of the subnets.
}

Our implementation does not build explicitly the proof trees for each
open transition, but instead : [Combining] enumerate all possible combinations of
pLTS transition (using TR1) and each synchronization vector choice
(TR2); [Matching] build the corresponding transition, and in particular
its predicate, in a hierarchical manner.
Each time we apply Tr2, we build a predicate as a conjunction of two
parts. The first part, which is the simpler one, composes the
predicate from the 
subnets and the guard from transitions in sub-pLTS or synchronization
vectors. The second part is the conjunction of equations generated
during matching the subnets' behaviors with the synchronization
vectors.

@@

\subsubsection{Result of LOTOS example}

@@

\QIN{
The final result open transitions represent all the
possible movements of the LOTOS example.
What can be observed is always the actions of the process $P$ from the beginning, until $P$ performs a $\delta(x)$ and the value of
$x$ is transmitted to the $acc(x);Q$ giving out a silent action. Such behavior is performed by the $ot_1$ and $ot_2$. 
Then it keeps presenting the behavior of $Q$. This can also be seen in $ot_3$. 
%The $ot_1$ transition means that the actions of the process $P$ could
%be transmitted out, until $P$ performs a $\delta(x)$ and the value of
%$x$ is transmitted, through transition
%$ot_2$. The other 3 open transitions shows the behaviors of the
%process $Q$ while the $ot_3$ and $ot_4$ won't be performed as the
%controller has already set the value $s_0$ to 1 in the previous
%action. 
}

@@

\QIN{
To sum up, from the pNets encoding the formulas of given algebra, which inputed by the user, we can exactly construct
the open automaton for them. 
}

@@

In this paper we presented the algorithm for generating an open
automaton representing the semantics of an open pNet, and we described
the implementation of the
algorithm. Our implementation includes two main parts. First, we
compute all the possible open transitions. The actions in pLTS and
hole behaviors are composed in the algorithm while the generation of
predicate need to match the combinations of the subnets with the
synchronization vectors. Some of the open transitions obtained at this
intermediate step, constructed in a structural manner from all
possible combinations of possible logical predicates, may
contain predicates which do not represent any possible concrete
instantiations. 
To get rid of these useless transitions, we use a SMT solver Z3 for checking the
satisfiability of the predicate in each open transition. In
order to do that, we have done some translation on the presentation,
the predicate and the assignments of the variables which are also
considered as a part of predicates before submitting them to the
Z3 solver. We implemented our algorithm in the VerCors platform and use some
running examples, the formula from other process algebras, to test the
algorithm. The result of the running examples are shown as the open
automaton with all the parameters generated by the algorithm
automatically. The open automaton successfully presents all the
possible movements of the LOTOS expression.

@@

Naturally, the next work after the generation of the open automata
will be to check the bisimulation of the pNets. In previous paper, we
have already found the FH-bisimulation, inspired by~\cite{deSimone85} as a
suitable definition. Usual approaches of bisimulation by partition
refinement does not work easily with symbolic transitions. In a first
step we may want instead to define and implement an algorithm fo
r checking that a given equivalence relation between open automata
states is indeed a FH-Bisimulation. Finding automatically a
bisimulation relation or the most general bisimulation will be more
challenging. It appears also that when comparing different pNets with
different structure, strong bisimulation will be too strong a notion,
and that we will have to define proper notions of weak equivalence or
refinements.

Before that work, we need to reconsider the result open transitions
from our algorithm. The ideal
result that we showed in the bottom part of
Fig.~\ref{schema:lotos-result} is not the one generated by the algorithm,
because it contains many intermediate
variables, such as fresh local variables of synchronization
vectors, and fresh result variables of intermediate open transitions.
Such intermediate variables come from the way predicates are
generated, and depends on the particular structure of the pNet. But
they are not significant in the resulting behavior, and should be
eliminated before comparing such behavior with another pNet.
To eliminate the intermediate variables in the predicate, we want to define
the set of the significant variables that we won't eliminate: the
local variables of pLTS transitions, the hole behaviors and the result
action at the top level.
The result of the elimination will be simpler predicates, that can be
successfully compared (e.g. by Z3) within the bisimulation checking
algorithm.

>>>>

We use an example inspired from LOTOS, to illustrate
the main features of pNets and our semantic construction algorithm.
%% \QIN{
%% Meanwhile, 
%% the formula using operators from CCS can also be encoded into a pNet and 
%% conducted with the same things.
%% }
%% \subsection{LOTOS Formula Example}
We choose here the expression \texttt{P>>(acc(v);Q)}, that includes the
Enable and Prefix 
operators of the LOTOS language. For the readers not familiar with
Lotos, the pNet encodings here should be sufficient to understand the
Enable operator semantics.
In Fig.~\ref{schema:lotos-pnet}, we show the encoding of the formula using pNets. 
This pNet is composed of: a toplevel node PN1 encoding the Enable
operator, a pLTS C1 acting as the {\sl controller} of the node; a hole
representing the 
process variable P; another node PN2 representing the Prefix
operator, with its controller C2, and its hole Q.

@@

% \TODO{Relate to the algorithm, or even move to section 5}
When building an Open Automaton as the semantics of a pNet, its
\emph{states}, and the shape of the \emph{predicates} in its
transitions have a specific structure:

\paragraph{States of open pNets:}\label{def-states}
  A state of an open pNet is a tuple of the
  states of its leaves (in which we denote tuples
  in structured states as $\triangleleft\ldots\triangleright$).
  For any pNet p, let $\overline{Leaves} = \mylangle S_i,{s_i}_0, \to_i\myrangle^{i \in L}$ be the set of pLTS at its leaves,
  then $States(p) = \{\triangleleft s_i^{i\in L}
  \triangleright| \forall i\in L. s_i \in S_i\}$.
A pLTS being its own single leave:
  $States(\mylangle S,s_0, \to\myrangle) = \{\triangleleft s \triangleright| s \in S\}$.
The initial state is defined as:
$InitState(p) = \triangleleft {{s_i}_0}^{i\in L}  \triangleright$.

@@

\paragraph{Predicates:}
Let
$\mylangle\overline{\pNet},\overline{S},\symb{SV}_k^{k\in K} \myrangle$
be a pNet. Consider a synchronization vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\[\MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Leftrightarrow
\begin{array}{l}
\exists {(a'_i)}^{i\in I},
{(b'_j)}^{j\in J},v'.\, SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
\\~~\land
\forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{array}\]
%% In any other case (if the action families do not match or if there is no valuation of
%% variables such that the above formula can be ensured) the predicate is undefined.

@@

\begin{figure}[h]
%  \vspace{-10mm}

@@

\section{Implementation}
\label{section:implementation}
The VerCors platform uses pNets as the intermediate language
for some high-level language or graphical formalism to be translated
into both input for a model-checker and for generating executable code
automatically \cite{HKM-FASE16}.
We have extended the pNet API in VerCors to deal with open pNets, and
also to specify the structure of action algebras.

In this section we describe the algorithm implementing the pNet
semantics, the interaction with the Z3 SMT solver, and we show the
result on our  example.

@@

\State Initialize sets $U$, $E$ for unexplored/explored global states
$U$, $L$ for result OTs;
\While{!isEmpty(U)}
	\State Chose a global state $S$ in $U$; recursively applying Tr1 or Tr2 on the root node;
	\If{Subnet is a pLTS}
                \State $s$ is the local state of the pLTS in the global state $S$;
		\For{each transition $tr$ starting from $s$}
			\State apply Tr1 to build the corresponding
                        $OT$; store $OT$ in $L$; \EndFor
	\EndIf
	\If{Subnet is a pNet, apply Tr2:} 
		\For{each $sv \in SV$}
			\State Invoke $Combining()$ to generate the
                        combination of sub-transitions, $Matching()$
                        to match the combination with the $sv$; then
                        store the result OTs in    $L$; 
		\EndFor
	\EndIf
	\For{each target global state $T$ from $OT \in L$}
		\IfThen{$(!U.contains(T))\ \&\&\ (!E.contains(T))$}{Add $T$ into the $U$;}
	\EndFor
\EndWhile
\State Prune the open transitions in $L$ using the SMT solver;
\State \Return $L$;

@@

Algorithm 1 shows the main structure of the algorithm: 
the computation repeats applying Tr1 and Tr2 to build all possible
combinations of behaviors of pLTS and Holes, through all possible
synchronisation vectors. The resulting open transitions are checked
for satisfiability by the SMT solver.

%From these, we compute
%the open automaton of an open pNet. We explain here our main
%implementation choices.
%:
%\begin{itemize}
%  \item Steps 2a-2b are merged, applying Tr2 from the
%premisses OTs from subnets to generate OTs without
%explicitely constructing the deduction tree.
%  \item We define a general naming schema for implementing the fresh 
%and clone functions with more structure than a simple global name
%generator. It ensures unicity, but builds names
%readable enough for debugging purpose.
%  \item Besides that, we also have designed a management of state variables and
%    assignments throughout the whole computation.
%%%   \item Satisfiability check is done only at
%%% top level of the open-transition construction. An alternative would be
%%% to submit the satisfiability check to the SMT solver at each level of
%%% the hierarchical OT construction, potentially reducing the overall number of
%%% combinations. But the submission to the SMT engine is costly, and more
%%% complexity analysis is required before deciding if this would be
%%% worthwhile. 
%\end{itemize}
%
%In the following sections, we detail:
%\begin{itemize}
%\item The structure of fresh variable names.
%\item The details of the implementation for step 2a-2b, refined into:
%  \subitem combining all possibilities of subnets activity/inactivity,
%  \subitem matching each combination with synchronization vectors.
%\item Management of local variables of pLTSs.
%\item Pruning unsatisfiable OTs using the Z3 SMT solver.
%\end{itemize}

@@

% \TODO{Xudong: move here a (new version) of the alogorithm loplevel sketch}
% \QIN{[Move to the head of Section 5.]}

%% \QIN{
%% The detail about the application of the Tr1 and Tr2 have not considered in the sketch, but it is an  
%% important part in the implementation of open transition generation. 
%% Each time we apply Tr1, there is not too much predicate terms, so do the other parameters, needed 
%% to be generated. 
%% When it comes to applying Tr2, the computation becomes complicated because it doing some composition of the subnets.
%% }

%\paragraph{Combining and Matching using TR2}
Our implementation does not build explicitly the proof trees for each
open transition, but instead : [Combining] enumerate all possible combinations of
pLTS transition (using TR1) and each synchronization vector choice
(TR2); [Matching] build the corresponding transition, and in particular
its predicate, in a hierarchical manner.
Each time we apply Tr2, we build a predicate as a conjunction of two
parts. The first part, which is the simpler one, composes the
predicate from the 
subnets and the guard from transitions in sub-pLTS or synchronization
vectors. The second part is the conjunction of equations generated
during matching the subnets' behaviors with the synchronization
vectors.

@@

% \subsubsection{Result of LOTOS example}

@@

The resulting open transitions represent all the
possible movements of the LOTOS example.
What can be observed is any actions of the process $P$ from the beginning, until $P$ performs a $\delta(x)$ and the value of
$x$ is transmitted to the $acc(x);Q$ giving out a silent action. Such behavior is performed by the $ot_1$ then $ot_2$. 
Then it keeps presenting the behavior of $Q$, as seen in $ot_3$. 
%The $ot_1$ transition means that the actions of the process $P$ could
%be transmitted out, until $P$ performs a $\delta(x)$ and the value of
%$x$ is transmitted, through transition
%$ot_2$. The other 3 open transitions shows the behaviors of the
%process $Q$ while the $ot_3$ and $ot_4$ won't be performed as the
%controller has already set the value $s_0$ to 1 in the previous
%action.

@@

%% \QIN{
%% To sum up, from the pNets encoding the formulas of given algebra, which inputed by the user, we can exactly construct
%% the open automaton for them. 
%% }

@@

In this paper we presented the algorithm for generating an open
automaton representing the semantics of an open pNet, and we described
the implementation of the
algorithm. Our implementation includes two main parts. First, we
compute all the possible open transitions. The actions in pLTS and
hole behaviors are composed in the algorithm while the generation of
predicate need to match the combinations of the subnets with the
synchronization vectors. Some of the open transitions obtained at this
intermediate step, constructed in a structural manner from all
possible combinations of possible logical predicates, may
contain predicates which do not represent any possible concrete
instantiations. 
To get rid of these useless transitions, we use the SMT solver Z3 for checking the
satisfiability of the predicate in each open transition. In
order to do that, we encode into Z3 the action algebra presentation,
the a representation of the predicates, before submitting them to the
Z3 solver. We implemented our algorithm in the VerCors platform and use some
running examples, encoding expressions from various process algebras, to test the
algorithm. Our use-cases show that our encoding
identifies successfully all unsatisfiable open transitions and that
the resulting automaton reflects correctly the expected 
movements of the encoded process expressions.

@@

Naturally, our next goals after the generation of the open automata
will be to model-check, and to check equivalence of  pNets. While
model-checking open automata (with a suitable logic including
predicates on data) seems easy to define, equivalence checking is more challenging.
In previous paper, we
have already found the FH-bisimulation, inspired by~\cite{deSimone85} as a
suitable definition. But usual approaches of bisimulation by partition
refinement does not work easily with symbolic transitions. In a first
step we may want instead to define and implement an algorithm for
checking that a given equivalence relation between open automata 
states is indeed a FH-Bisimulation. Finding automatically a
bisimulation relation or the most general bisimulation will be more
challenging. It appears also that when comparing different pNets with
different structure, strong bisimulation will be too strong a notion,
and that we will have to define proper notions of weak equivalence or
refinements.

Before that work, we need to reconsider the result open transitions
from our algorithm. The ideal
result that we showed in the bottom part of
Fig.~\ref{schema:lotos-result} is not the one generated by the algorithm,
because the original one contains many intermediate
variables, such as fresh local variables of synchronization
vectors, and fresh result variables of intermediate open transitions.
Such intermediate variables come from the way predicates are
generated, and depends on the particular structure of the pNet. But
they are not significant in the resulting behavior, and should be
eliminated before comparing such behavior with another pNet.
To eliminate the intermediate variables in the predicate, we want to define
the set of the significant variables that we won't eliminate: the
local variables of pLTS transitions, the hole behaviors and the result
action at the top level.
The result of the elimination will be simpler predicates, that can be
successfully compared (e.g. by Z3) within the bisimulation checking
algorithm.
